---
title: "R Notebook"
output: 
  html_document:
    toc: true
    toc-float: true
    self_contained: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.height = 6, fig.width = 10, fig.retina = 2,
  warning = FALSE, message = FALSE
)
library(tidyverse)
```

## Background

[The Association Between Income and Life Expectancy in the United States, 2001-2014](https://jamanetwork.com/journals/jama/fullarticle/2513561)

## Load Data

```{r load}
life_exp_file <- here::here("data", "health_ineq_online_table_9.csv")
life_exp <- read_csv(life_exp_file)
```

## Tidy Data

```{r tidy}
income_quantiles <- c(
  "1" = "Bottom", 
  "2" = "Lower Middle", 
  "3" = "Upper Middle", 
  "4" = "Top"
)
adjustment_types <- c(raceadj = "Race Adjusted", agg = "Unadjusted")

life_exp_tidy <-
  life_exp %>% 
  select(cz:year, contains("le_")) %>% 
  gather(key, value, contains("le_")) %>% 
  extract(
    col = key, 
    into = c("variable", "adjustment", "income_quantile", "sex"), 
    regex = "(sd_le|le)_(raceadj|agg)_q(\\d)_(\\w)"
  ) %>% 
  mutate(
    adjustment = recode(adjustment, !!!adjustment_types),
    income_quantile = recode(income_quantile, !!!income_quantiles),
    sex = recode(sex, "F" = "Female", "M" = "Male")
  ) %>% 
  rename(state = stateabbrv) %>% 
  spread(variable, value)
```

## Exploratory Plot: Life Expectancy by Gender and Income Quantile in Tampa

```{r plot-tampa-life-expectancy-gender}
income_quantile_colors <- c(
  "Bottom"       = "#00589a",
  "Lower Middle" = "#82c878",
  "Upper Middle" = "#faa555",
  "Top"          = "#eb1455"
)

life_exp_tidy %>% 
  filter(
    czname == "Tampa",
    adjustment == "Race Adjusted",
    income_quantile %in% c("Bottom", "Top")
  ) %>% 
  ggplot() +
  aes(
    x = year, y = le, 
    color = income_quantile, 
    group = paste(income_quantile, sex)
  ) +
  geom_point(size = 0.75, show.legend = FALSE) +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ sex) +
  scale_color_manual(values = income_quantile_colors) +
  scale_x_continuous(breaks = seq(2002, 2018, 4)) +
  labs(
    title = "Tampa",
    x = NULL,
    y = "Life Expectancy",
    color = "Income Quantile"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Intermediate Processing: Calculate Mean (Male, Female) Life Expectancy

```{r calculate-mean-life-expectancy}
life_exp_tidy_mean <- 
  life_exp_tidy %>% 
  gather(variable, value, le:sd_le) %>% 
  nest(sex, value) %>% 
  mutate(value = map_dbl(data, ~ mean(.$value))) %>% 
  select(-data) %>% 
  spread(variable, value)
```

## Exploratory Plot: Mean Life Expectancy by Income Quantile in Tampa

```{r plot-tampa-mean-life-expectancy}
life_exp_tidy_mean %>% 
  filter(
    czname == "Tampa",
    adjustment == "Race Adjusted"
  ) %>% 
  ggplot() +
  aes(x = year, y = le, color = income_quantile) +
  geom_point(show.legend = FALSE) +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(breaks = seq(2002, 2018, 4)) +
  theme_minimal() +
  labs(
    title = "Tampa",
    x = NULL,
    y = "Life Expectancy",
    color = "Income Quantile"
  ) +
  scale_color_manual(values = income_quantile_colors) +
  theme(legend.position = "bottom")
```

## Model: Life Expectancy by Income Quantile

```{r model-race-adjusted-life-expectancy}
life_exp_model <- 
  life_exp_tidy_mean %>% 
  select(czname, state, pop2000, year, adjustment, income_quantile, le) %>% 
  filter(
    adjustment == "Race Adjusted",
    income_quantile %in% c("Bottom", "Top")
  ) %>% 
  nest(year, le) %>% 
  mutate(
    lm = map(data, ~ lm(le ~ year, data = .x)),
    pred = map2(data, lm, modelr::add_predictions),
    est = map(lm, broom::tidy)
  )

head(life_exp_model)
```

## Report: Extract Results from Model

```{r model-results-race-adjusted-life-expectancy}
life_exp_model_results <- 
  life_exp_model %>% 
  unnest(est) %>% 
  filter(term == "year") %>% 
  select(-adjustment, -term, -std.error, -statistic, -p.value) %>% 
  spread(income_quantile, estimate) %>% 
  mutate(czname = fct_reorder(czname, Bottom, .desc = TRUE)) %>% 
  arrange(czname) %>% 
  rename(
    "Commuting Zone" = czname,
    "State" = state,
    "Population (2000)" = pop2000
  )
```

```{r model-table-race-adjusted-life-expectancy}
life_exp_model_results %>% 
  head(n = 10) %>% 
  knitr::kable(digits = 3, caption = "Top 10 Cities")

life_exp_model_results %>% 
  tail(n = 10) %>% 
  knitr::kable(digits = 3, caption = "Worst 10 Cities")
```

## Report: Plot Modeled Life Expectancy in Best and Worst Cities

```{r model-rank-race-adjusted-life-expectancy}
plot_modeled_life_exp <- function(income_quantile, group, data) {
  data %>% 
    arrange(rank) %>% 
    mutate(
      city = paste(czname, state, sep = ", "),
      city = fct_inorder(city)
    ) %>% 
    ggplot() +
    aes(year, pred, color = city) +
    geom_line() +
    scale_y_continuous(limits = c(75, 90)) +
    scale_x_continuous(breaks = seq(2002, 2018, 4)) +
    labs(x = NULL, y = "Life Expectancy", color = NULL) +
    ggtitle(
      if (group == "Best") glue::glue("{income_quantile} Income Quantile"),
      glue::glue("{group} Life Expectancy Change")
    ) +
    theme_minimal()
}

life_exp_model %>% 
  mutate(
    coef = map_dbl(est, ~ .x %>% filter(term == "year") %>% pull(estimate))
  ) %>% 
  group_by(income_quantile) %>% 
  mutate(rank = min_rank(desc(coef))) %>% 
  ungroup() %>% 
  filter(!between(rank, 7, 94)) %>% 
  mutate(group = case_when(rank <= 10 ~ "Best", TRUE ~ "Worst")) %>% 
  arrange(rank) %>% 
  unnest(pred) %>% 
  nest(-income_quantile, -group) %>% 
  arrange(group, income_quantile) %>% 
  mutate(plot = pmap(., plot_modeled_life_exp)) %>% 
  pull(plot) %>% 
  cowplot::plot_grid(plotlist = ., ncol = 2)
```
